preciseTAD: A machine learning framework for precise TAD boundary prediction at base pair resolution

### 2020-07-01 Manuscript/manuscript.Rmd
cleaned up manuscript by adding revisions; changed title and abstract; changed "Comparing called vs predicted boundaries" section to deeptools; added refs

---
title: Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings
csl: styles.ref/genomebiology.csl
output:
  pdf_document: default
  word_document:
    reference_docx: styles.doc/NIH_grant_style.docx
bibliography: References/references.bib
---

@@

**Background.** Topologically Associated Domains (TADs) represent fundamental building blocks that lead to the organization and regulation of the three-dimensional (3D) genome. The enrichment of several epigenetic marks at TAD boundaries suggests a strong epigenetic component of TAD boundary formation. Several methods have been developed to leverage the growing amount of (epi)genomic annotation data for predicting TAD boundaries; however, they overlooked key characteristics of the data including spatial associations between genomic annotations and boundaries, as well as the severe imbalance in TAD-boundary vs non-TAD boundary defined genomic regions. Furthermore, resolution of Hi-C data remains well below the resolution of epigenomic assays (hundredths of bases for ChIP-seq) affecting the ability of Hi-C sequencing in discerning gene-level features of chromatin. This resolution gap between 1D and 3D genomic maps, as well as a paucity of chromatin features below the level of TADs, has significantly limited our understanding of epigenomic regulation of gene expression and its potential link to chromatin architecture.

**Methods.** We first developed an ensemble framework for optimizing data level characteristics necessary for TAD boundary prediction including combinations of 5 Hi-C data resolutions (5 kb to 100 kb), 3 spatially defined predictor types, and 4 re-sampling techniques, across the GM12878 and K562 cell lines. We then aimed to predict TAD boundary location on a 1D genomic map by validating optimally built cell line specific models at basepair resolution.

**Results.** We demonstrate with our ensemble framework that cell line specific random forest classifiers built on 10 kb resolution Hi-C data, with distance-type predictors, using SMOTE (*Synthetic Minority Oversampling Technique*) re-sampling yield optimal predictive performances, and outperform existing methods. Cell specific transcription factor binding sites (TFBS) were the best type of genomic annotations for predicting TAD boundaries when directly compared with histone modifications and chromatin states; we identified the well-known CTCF, SMC3, RAD21, and ZNF143 TFBS known to be enriched at TAD boundaries. Additionally, TAD boundary regions predicted at basepair resolution using our optimally built cell line specific models were found to be significantly closer to known enriched epigenomic marks indicating more accurate identification of TAD boundaries than existing Hi-C sequencing technologies.

@@

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of important 3-dimensional (3D) spatial constructs within the human genome such as chromatin loops [@rao20143d], Topologically Associated Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At kilobase scale, gene promotors are often connected via distal regulatory elements, such as enhancers, through chromatin loops, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. Further out, at the megabase scale, TADs represent regions on the linear genome, that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation. It has been shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological;@krefting2018evolutionary]. Likewise, disruption of the boundaries that demarcate TADs and loops have been shown to lead to cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Therefore, precisely identifying the location of TAD and loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

Several methods have been proposed to identify the genomic coordinates of TAD boundaries throughout the linear genome [@dixon2012topological;@rao20143d;@sexton2012three;@chen2016novo;@filippova2014identification;@shin2016topdom;@weinreb2016identification], less so for chromatin loops [@salameh2019supervised;@kai2018predicting]. [??? Check refs, the current ones are suboptimal. Check SpectralTAD paper for refs about TAD callers, use https://github.com/mdozmorov/HiC_tools#loop-callers for loop callers] However, a key limitation among most of these TAD- and loop callers, referred hereafter as TAD callers for brevity, is that they are heavily reliant on Hi-C data *resolution*. Resolution refers to the sizes of bins used to segment the linear genome as a way to deal with low signal-to-noise incurred by the high dimensional nature of Hi-C data [@li2019comparison]. Lower resolutions contribute to larger bin sizes, which lead to uncertainty in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TADs and chromatin loops [@fudenberg2016formation].

Another limitation among TAD callers is that they disregard prior knowledge about functional genomic elements associated with TAD and loop boundaries. The insulator binding protein, CTCF, and additional cofactors such as SMC3 and RAD21 have been hypothesized to comprise components of the loop extrusion model, whereby DNA is extruded through cohesin rings forming loops [@hansen2018recent;@fudenberg2016formation;@alipour2012self;@sanborn2015chromatin]. [??? Rewrite, simplify, describe enrichment first, loop extrusion second] Additionally, distinct patterns of histone modifications have also been shown to be featured at boundaries, including H3K36 timethylation (H3K36me3) [@lieberman2009comprehensive;@dixon2012topological;@libbrecht2015joint]. These elements are annotated on the genome using chromatin immunoprecipitation followed by high-throughput DNA sequencing (ChIP-seq). Even at its highest (750 bp [@bonev2017multiscale]), Hi-C sequencing resolution still remains well below the resolution of ChIP-seq experiments, which ranges in the tens of base pairs [@mendoza2009genome]. Therefore, leveraging such elements in a supervised learning framework lends itself to the possibility of more precisely predicting the locations of TAD (or loop) boundaries at higher resolutions.

Here we present *preciseTAD*, an R package for precise identification of TAD(or loop) under a supervised machine learning framework. Our method builds upon an optimized random forest classification model (RF) that leverages high resolution ChIP-seq data in the form of regulatory elements (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS). We introduce a novel technique for building RF classifiers in a boundary prediction setting composed of a resolution-specific binning process and feature engineering procedure. *preciseTAD* then employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) on prediction probabilities to select the most biologically meaningful boundary coordinates as base pairs on the linear genome, thereby circumventing resolution restrictions of conventional TAD- and loop-callers. We demonstrate that predictions made by *preciseTAD* are more enriched for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines, highlighting their biological significance. This improved precision in TAD and loop boundary location can provide insight into the role that epigenomics has on the spatial architecture of the human genome. [??? Rewrite. Say the main methodology first - model on 4 TFBSs, uses distance, RUS, technique for bp-level prediction, one sentence per point. Detail next - BroadHMM and histone were tested, OC, OP were tested, various sampling techniques. Conclude with findings and package availability]

@@

In Hi-C, each chromosome is binned into non-overlapping regions of the same length *r*, typically 5kb and above. The *r* parameter defines the resolution of Hi-C data. Consequently, the location of TAD boundaries is limited by the resolution *r*. We designed a strategy called _shifted binning_ that partitions the genome into regions of the same length *r*, but with middle points corresponding to boundaries defined by the original binning.

@@

Cell line-specific genomic annotations were used to build the feature space for our boundary prediction framework. Bins were annotated by one of either the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average signal strength of the corresponding annotation (*Peak Signal Strength*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*). A $log_2$-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). It should be noted that models built using a *Peak Signal Strength* feature space were only composed of histone modifications and transcription factor binding sites. That is, they did not include BroadHMM chromatin segmentations.

@@

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, feature engineering procedures, and re-sampling techniques. First, the data matrix $A_{N \times (p+1)}$ was constructed by combining the binned genome from each of the $n-1$ autosomal chromosomes ($1,2,\cdots, i-1,i+1,\cdots,22$) as well as a particular feature engineering procedure. Re-sampling was then performed on $A$. An RF classifier was then trained on $A$ using 5-fold cross-validation to tune for the number of annotations, *p*, to consider at each node (*mtry*). The number of trees (*ntree*) that were aggregated for each RF model was set to 500. The minimum number of observations per root node (*nodesize*) was set to 1. To evaluate model performance, the $i^{th}$ chromosome was used for testing. The performance of each combination of model inputs was compared using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

<!--
$$F1 - Score = \frac{2TP}{2TP + FP + FN}$$
-->

<!--
$$
MCC = \frac{TP \times TN - FP \times FN}{\sqrt{ (TP + FP)(TP+FN)(TN+FP)(TN+FN) }}
$$
-->

@@

<!--
$$Informedness = \dfrac{TP}{TP+FN} + \dfrac{TN}{TN+FP} - 1$$
-->

where TP refers to the number of bins correctly identified as containing a boundary (true positives), FP refers to the number of bins incorrectly identified as containing a boundary (false positives), TN refers to the number of bins correctly identified as not containing a boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a boundary (false negatives). Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. <!--We refer to the complete computational framework used to compare random forest model inputs as *TADRF*.--> Performances were aggregated across each $i^{th}$ holdout chromosome using means and standard deviations.

@@

Many genomic annotations, notably architectural proteins, tend to exhibit an extensive pattern of colocalization (correlation) [@van2014insulator]. To suitably reduce the feature space and improve computational efficiency, while maintaining optimal performance, we implemented recursive feature elimination. In doing so, we estimated the near optimal number of necessary features to include in the best fitting RF model. The best performing subset was identified, ranging from 2 to the maximum number of features by power of 2. We then aggregated the predictive importance of the union of the optimal set of features across chromosomes using the mean decrease in node impurity among permuted features in out-of-bag samples to determine the most common and top ranked annotations for predicting boundary regions.

@@

To investigate whether we could alleviate resolution limitations of conventional TAD and chromatin loop calling tools, we developed *preciseTAD*, an algorithm that leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base pair (bp) resolution. To demonstrate, we implemented an optimally built RF model using the framework described above. That is, a random forest model built on the optimal combination of data resolution (binning), predictor type (feature space), re-sampling technique (class balancing correction), and annotations (most informative genomic elements).

First the vector, $p_{n_{i}}$, denoting each base pair's probability of being a boundary point from the RF model, was extracted, where $n_{i}$ represents the length of chromosome $i$. Next, *DBSCAN* (version 1.1-5) was applied to the matrix of pairwise genomic distances between base pairs with $p_{n_{i}}=1$. This was to avoid the inflation of predicted boundaries that can occur naturally when dealing with such fine resolution, namely, due to spatially close base pairs with similar predictor profiles (i.e. similar overlap counts, distances, etc.). The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts*. The eps parameter was assigned as the respective data resolution. The heuristic being that density-reachable base pairs with genomic distances less than resolution should occupy the same designated cluster. The corresponding MinPts parameter was set to recommended value of 3, representing $1+dim(data)$ [@hahsler2019dbscan]. The resulting clusters of highly predictive base pairs identified by DBSCAN were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base pair among each PTBR, *preciseTAD* then implements an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. Thusly, the corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative and biologically meaningful base pair within each clustered PTBR.

@@

Next, we evaluated the biological significance of our predicted boundaries with respect to the enrichment of several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. Enrichmed heatmaps were used to compare preciseTAD boundaries to both ARROWHEAD and PEAKACHU called boundaries using the `EnrichedHeatmap` R package (version 1.14.0) [@gu2018enrichedheatmap]. For each boundary (both called and predicted), a 5000 bp flank was created on either side. The flanking was broken up into 50 bp segments, for a total of 100 windows on both sides of a given boundary. A coverage value was generated for each window with respect to a given ChIP-seq annotation, as the mean signal averaged by the width of the window, denoted by,

$$v_{c} = \dfrac{\Sigma_{j}^{m} \Sigma_{i}^{n}x_{ij}w_{ij}}{L}$$
where $x_{ij}$ is the total number of base pairs of annotation $i$ in window $j$ and $w_{ij}$ is the number of base pairs that overlap between annotation $i$ and window $j$. The denominator, $L$, is the width of the windows (here, $L=50$). The rows of the heatmaps are ordered by enriched scores. The enriched scores are calculated as the sum of coverage values weighted by the distance to the flanked boundary, denoted by

$$s_{e} = \Sigma_{d=1}^{n_{1}}\left(\dfrac{x_{1d}* d}{n_{1}} \right) + \Sigma_{u=1}^{n_{2}}\left(\dfrac{x_{2u}* (n_{2} - u + 1)}{n_{2}} \right)$$
where $n_{1}$ and $n_{2}$ represent the number of downstream and upstream coverage values to sum over. Here, $n_{1}=n_{2}=100$.

Additionally, we compared the median $log_2$ distances between TAD boundaries and the same top predictive transcription factor binding sites using Wilcoxon Rank-Sum test.

Furthermore, under the hypothesis that TAD boundaries are highly concerved across cell types [@dixon2012topological; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa], we assessed the conservation of our predicted TAD boundaries across cell lines and prediction algorithms. Boundaries were flanked by resolution and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0) and evaluated using the Jaccard index defined as

@@

where A and B represent genomic regions created by flanking called and predicted boundaries. That is, between cell lines, the number of overlapping flanked boundaries divided by the total number of flanked boundaries. Wilcoxon rank sum tests were used to compare chromosome-specific jaccard indices across cell lines, between *preciseTAD* boundaries and both ARROWHEAD and PEAKACHU boundaries respectively. The alpha level of significance was set to 0.05 for all statistical tests.

@@

## Developing a computational framework for optimal TAD boundary prediction

We developed a computational framework<!--, refered to as *TADRF*,--> aimed at determining the optimal set of data level characteristics necessary for topologically associated domain (TAD) boundary prediction. We chose the random forest (RF) machine learning algorithm as our binary classification tool. The reason for choosing RF was two-fold: (1) In order to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors with respect to their ability to predict boundary regions [@boulesteix2012overview]. We then leveraged our optimally built RF to precisely predict the genomic coordinates of both TAD and loop boundaries at base pair resolution.

Our framework is resolution-specific and uses called TAD and loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). TADs were subsequently called using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loop boundaries we opted to use boundaries obtained by the PEAKACHU tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? TAD and loop boundaries should be distinguished, to describe when we have all results]

Boundary regions were defined as regions containing a called boundary ($Y = 1$), while non-boundary regions were defined as genomic bins that did not contain a boundary ($Y = 0$) (Figure 1A, see Methods). The total number of TADs, TAD boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 3). The number of non-boundary regions highly outnumbered boundary regions, and these results were consistent for loop boundaries (Supplementary Table 4). Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects model performance [@prati2011survey; @wei2013role]. To address class imbalance, we evaluated the effect of three re-sampling techniques on prediction performance. *Random over-sampling* (ROS) was defined as sampling with replacement from boundary regions. *Random under-sampling* (RUS) was defined as sampling with replacement from non-boundary regions. Lastly, we considered *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling [@chawla2002smote] (see Methods). These strategies allowed us to select the best performing model.

A total of 78 cell line specific functional genomic annotations were used to build the feature space of downstream random forest models. These annotations included BroadHMM shromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS, Supplementary Table 2). Four feature engineering procedures were developed for each genomic annotation (Figure 1B, see Methods) with the goal to improve boundary predictions by accounting for signal strength association (Signal), direct (OC), proportional (OP), and spatial (Distance) relationships between TAD boundaries and genomic annotations.

@@

Expectedly, when using data with class imbalance present, that is, the data were used as-is, the models exhibited low balanced accuracies, with very little variability among different resolutions (Figure 3). Similarly, poor performances were found when using ROS as well. However, RUS and SMOTE re-sampling led to drastic improvement in performance, especially at higher resolutions. We found that RUS marginally outperformed SMOTE as the most optimal class balancing technique for all resolutions and predictor types. These results suggest that simple randomly under-sampling alone can drastically improve TAD boundary prediction.

>>>>

---
title: "preciseTAD: A machine learning framework for precise TAD boundary prediction at base pair resolution"
csl: styles.ref/genomebiology.csl
output:
  word_document:
    reference_docx: styles.doc/NIH_grant_style.docx
  pdf_document: default
bibliography: References/references.bib
---

@@

Chromosome conformation capture combined with high-throughput sequencing experiments (Hi-C) have revealed that chromatin undergoes layers of compaction through DNA looping and folding, forming dynamic 3D structures. Among these are Topologically Associating Domains (TADs), which are known to play critical roles in cell dynamics like gene regulation and cell differentiation. Precise TAD mapping remains difficult, as it is strongly reliant on Hi-C data resolution. Obtaining genome-wide chromatin interactions at high-resolution is costly resulting in variability in true TAD boundary location by TAD calling algorithms. To aid in the precise identification of TAD boundaries we developed preciseTAD, a machine learning framework that leverages the spatial relationship of many high resolution ChIP-seq defined genomic elements, coupled with density-based clustering  and scalable partitioning techniques. Our framework precisely predicts chromosome-specific TAD boundaries on multiple cell types. We show that known molecular drivers of 3D chromatin including CTCF, RAD21, and SMC3 are more enriched at our predicted TAD boundaries compared to the boundaries identified by the popular ARROWHEAD TAD caller. Our results provide useful insights into the 3D organization of the human genome.

@@

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of important 3-dimensional (3D) spatial constructs within the human genome such as chromatin loops [@rao20143d], Topologically Associated Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At kilobase scale, gene promotors are often connected via distal regulatory elements, such as enhancers, through chromatin loops, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. Further out, at the megabase scale, TADs represent regions on the linear genome, that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological;@krefting2018evolutionary]. Likewise, disruption of the boundaries that demarcate TADs and loops have been shown to lead to cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Therefore, precisely identifying the location of TAD and chromatin loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

Several methods have been proposed to identify the genomic coordinates of TAD boundaries throughout the linear genome [@dixon2012topological;@rao20143d;@sexton2012three;@chen2016novo;@filippova2014identification;@shin2016topdom;@weinreb2016identification], less so for chromatin loops [@salameh2019supervised;@kai2018predicting]. [??? Check refs, the current ones are suboptimal. Check SpectralTAD paper for refs about TAD callers, use https://github.com/mdozmorov/HiC_tools#loop-callers for loop callers] However, a key limitation among most of these TAD- and loop callers, referred hereafter as TAD callers for brevity, is that they are heavily reliant on Hi-C data *resolution*. Resolution refers to the sizes of genomic bins used to segment the linear genome as a way to deal with low signal-to-noise incurred by the high dimensional nature of Hi-C data [@li2019comparison]. Lower resolutions contribute to larger bin sizes, which lead to uncertainty in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TADs and chromatin loops [@fudenberg2016formation].

Another limitation among TAD callers is that they disregard prior knowledge about functional genomic elements associated with TAD and loop boundaries. The insulator binding protein, CTCF, and additional cofactors such as SMC3 and RAD21 have been hypothesized to comprise components of the loop extrusion model, whereby DNA is extruded through cohesin rings forming chromatin loops [@hansen2018recent;@fudenberg2016formation;@alipour2012self;@sanborn2015chromatin]. [??? Rewrite, simplify, describe enrichment first, loop extrusion second] Additionally, distinct patterns of histone modifications have also been shown to be featured at boundaries, including H3K36 timethylation (H3K36me3) [@lieberman2009comprehensive;@dixon2012topological;@libbrecht2015joint]. These elements are annotated on the genome using chromatin immunoprecipitation followed by high-throughput DNA sequencing (ChIP-seq). Even at its highest (750 bp [@bonev2017multiscale]), Hi-C sequencing resolution still remains well below the resolution of ChIP-seq experiments, which ranges in the tens of base pairs [@mendoza2009genome]. Therefore, leveraging such elements in a supervised machine learning framework lends itself to the possibility of more precisely predicting the locations of TAD (or loop) boundaries at much higher resolutions.

Here we present *preciseTAD*, an R package for precise identification of TAD(or loop) boundaries under a supervised machine learning framework. Our method builds upon an optimized random forest (RF) classification model that leverages high resolution ChIP-seq data in the form of regulatory elements (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS). We introduce a novel technique for building RF classifiers in a boundary prediction setting composed of a resolution-specific binning process and feature engineering procedure. *preciseTAD* then employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) on prediction probabilities to select the most biologically meaningful boundary coordinates on the linear genome, thereby circumventing resolution restrictions of conventional TAD- and chromatin loop-callers. We demonstrate that predictions made by *preciseTAD* are more enriched for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines, highlighting their biological significance. This improved precision in TAD and loop boundary location can provide insight into the role that epigenomics has on the spatial architecture of the human genome. [??? Rewrite. Say the main methodology first - model on 4 TFBSs, uses distance, RUS, technique for bp-level prediction, one sentence per point. Detail next - BroadHMM and histone were tested, OC, OP were tested, various sampling techniques. Conclude with findings and package availability]

@@

In Hi-C, each chromosome is binned into non-overlapping regions of the same length *r*, typically 5 kb and above. The *r* parameter defines the resolution of Hi-C data. Consequently, the location of TAD boundaries is limited by the resolution *r*. Here, we designed a strategy called _shifted binning_ that partitions the genome into regions of the same length *r*, but with middle points corresponding to boundaries defined by the original binning.

@@

Cell line-specific genomic annotations were used to build the feature space for our boundary prediction framework. Bins were annotated by one of either the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average signal strength of the corresponding annotation (*Peak Signal Strength*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*). A $log_2$-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). It should be noted that models built using a *Peak Signal Strength* feature space were only composed of histone modifications and transcription factor binding sites. That is, they did not include any BroadHMM chromatin segmentations.

@@

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, feature engineering procedures, and re-sampling techniques. First, the data matrix $A_{N \times (p+1)}$ was constructed by combining the binned genome from each of the $n-1$ autosomal chromosomes ($1,2,\cdots, i-1,i+1,\cdots,22$) as well as a particular feature engineering procedure. Re-sampling was then performed on $A$. An RF classifier was then trained on $A$ using 3-fold cross-validation to tune for the number of annotations, *p*, to consider at each node (*mtry*). The number of trees (*ntree*) that were aggregated for each RF model was set to 500. The minimum number of observations per root node (*nodesize*) was set to 1. To evaluate model performance, the $i^{th}$ chromosome was used for testing. The performance of each combination of model inputs was compared using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

@@

where TP refers to the number of bins correctly identified as containing a boundary (true positives), FP refers to the number of bins incorrectly identified as containing a boundary (false positives), TN refers to the number of bins correctly identified as not containing a boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a boundary (false negatives). Each of these quantities are obtained from the confusion matrix created by validating the model on the test data. Performances were aggregated across each $i^{th}$ holdout chromosome using means and standard deviations.

@@

Many genomic annotations, notably architectural proteins, tend to exhibit an extensive pattern of colocalization (correlation) [@van2014insulator]. To suitably reduce the feature space and improve computational efficiency, while maintaining optimal performance, we implemented recursive feature elimination (RFE). In doing so, we estimated the near optimal number of necessary features to include in the best fitting RF model. The best performing subset was identified, ranging from 2 to the maximum number of features by powers of 2. We then aggregated the predictive importance of the union of the optimal set of features across chromosomes using the mean decrease in node impurity among permuted features in out-of-bag samples to determine the most common and top ranked annotations for predicting boundary regions.

@@

To investigate whether we could alleviate resolution limitations of conventional TAD and chromatin loop calling tools, we developed *preciseTAD*, an algorithm that leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base pair (bp) resolution. To demonstrate, we implemented an optimally built RF model using the framework described above. That is, a random forest model built on the optimal combination of data resolution (binning), predictor type (feature space), re-sampling technique (class balancing correction), and top ranked annotations (most informative genomic elements).

First the vector, $p_{n_{i}}$, denoting each base pair's probability of being a boundary point from the RF model, was extracted, where $n_{i}$ represents the length of chromosome $i$. Next, *DBSCAN* (version 1.1-5) was applied to the matrix of pairwise genomic distances between base pairs with $p_{n_{i}}=1$. This was to avoid the inflation of predicted boundaries that can occur naturally when dealing with such fine resolution, namely, due to spatially close base pairs with similar predictor profiles (i.e. similar overlap counts, distances, etc.). The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts*. The eps parameter was assigned as the respective data resolution. The heuristic being that density-reachable base pairs with genomic distances less than resolution should occupy the same designated cluster. The corresponding MinPts parameter was set to recommended value of 3, representing $1+dim(data)$ [@hahsler2019dbscan]. The resulting clusters of highly predictive base pairs identified by DBSCAN were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base pair among each PTBR, *preciseTAD* then implements an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. Thusly, the corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative and biologically meaningful base pair coordinate within each clustered PTBR.

@@

afgan2018galaxy, ramirez2016deeptools2

Next, we assessed the biological significance of our predicted boundaries with respect to the enrichment profiles of several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the computeMatrix and plotHeatmap tools offered by DeepTools [@ramirez2016deeptools2] via integration with the Galaxy project [@afgan2018galaxy]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum test.

Furthermore, we assessed the hypothesis that TADs and chromatin loops are conserved  [@dixon2012topological; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa] by comparing predicted and called boundaries between the GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0) and evaluated using the Jaccard index defined as

@@

where A and B represent genomic regions created by flanking called and predicted boundaries. That is, between cell lines, the number of overlapping flanked boundaries divided by the total number of flanked boundaries. Wilcoxon rank sum tests were used to compare chromosome-specific jaccard indices across cell lines, between *preciseTAD* boundaries and both ARROWHEAD and PEAKACHU boundaries respectively. All statistical analyses were performed in R version 4.0.1. The alpha level of significance was set to 0.05 for all statistical tests.

<!--using the `EnrichedHeatmap` R package (version 1.14.0) [@gu2018enrichedheatmap]. For each boundary (both called and predicted), a 5000 bp flank was created on either side. The flanking was broken up into 50 bp segments, for a total of 100 windows on both sides of a given boundary. A coverage value was generated for each window with respect to a given ChIP-seq annotation, as the mean signal averaged by the width of the window, denoted by,

$$v_{c} = \dfrac{\Sigma_{j}^{m} \Sigma_{i}^{n}x_{ij}w_{ij}}{L}$$

where $x_{ij}$ is the total number of base pairs of annotation $i$ in window $j$ and $w_{ij}$ is the number of base pairs that overlap between annotation $i$ and window $j$. The denominator, $L$, is the width of the windows (here, $L=50$). The rows of the heatmaps are ordered by enriched scores. The enriched scores are calculated as the sum of coverage values weighted by the distance to the flanked boundary, denoted by

$$s_{e} = \Sigma_{d=1}^{n_{1}}\left(\dfrac{x_{1d}* d}{n_{1}} \right) + \Sigma_{u=1}^{n_{2}}\left(\dfrac{x_{2u}* (n_{2} - u + 1)}{n_{2}} \right)$$
where $n_{1}$ and $n_{2}$ represent the number of downstream and upstream coverage values to sum over. Here, $n_{1}=n_{2}=100$.-->

@@

## Developing a machine learning framework for optimal TAD boundary prediction

We developed a machine learning framework<!--, refered to as *TADRF*,--> aimed at determining the optimal set of data level characteristics necessary for topologically associated domain (TAD) boundary prediction. We chose the random forest (RF) machine learning algorithm as our binary classification tool. The reason for choosing RF was two-fold: (1) In order to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors with respect to their ability to predict boundary regions [@boulesteix2012overview]. We then developed *preciseTAD*, an algorithm which leverages an optimally built RF to precisely predict the genomic coordinates of both TAD and chromatin loop boundaries at base pair resolution.

Our framework is resolution-specific and uses called TAD or loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines obtained from GEO GSE63525 (Supplementary Table 1). TADs were subsequently called using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loop boundaries, we opted to use boundaries obtained by the PEAKACHU prediction tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? TAD and loop boundaries should be distinguished, to describe when we have all results]

Boundary regions were defined as regions containing a called boundary ($Y = 1$), while non-boundary regions were defined as genomic bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of TADs, TAD boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 3). The number of non-boundary regions highly outnumbered boundary regions, with results consistent for loop boundaries as well (Supplementary Table 4). Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects model performance [@prati2011survey; @wei2013role]. To address class imbalance, we evaluated the effect of three re-sampling techniques on prediction performance. *Random over-sampling* (ROS) was defined as sampling with replacement from boundary regions. *Random under-sampling* (RUS) was defined as sampling with replacement from non-boundary regions. Lastly, we considered *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling [@chawla2002smote] (see Methods). These strategies allowed us to select the best performing model.

A total of 78 cell line specific functional genomic annotations were used to build the feature space of downstream random forest models. These annotations included BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS, Supplementary Table 2). Four feature engineering procedures were developed for each genomic annotation (Figure 1B, see Methods) with the goal to improve boundary predictions by accounting for signal strength association (Signal), direct (OC), proportional (OP), and spatial (Distance) relationships between TAD boundaries and genomic annotations.

@@

Expectedly, when using data with class imbalance present, that is, no re-sampling, the models exhibited low balanced accuracies, with very little variability among different resolutions (Figure 3). Similarly, poor performances were found when using ROS as well. However, RUS and SMOTE re-sampling led to drastic improvement in performance, especially at higher resolutions. We found that RUS marginally outperformed SMOTE as the most optimal class balancing technique for all resolutions and predictor types. These results suggest that simple random under-sampling alone can drastically improve TAD boundary region prediction.
